---
title: "A Tutorial of GPCCA Implementation to a Multi-modal Dataset"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{A Tutorial of GPCCA Implementation to a Multi-modal Dataset}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Preparation

Suppose the R package `GPCCA` is correctly installed. We attach it to the environment.

```{r}
library(GPCCA)
```

Firstly, we briefly introduce the example multi-modal dataset provided by the package. 
`example_MultiModalData_list` is a synthetic multi-modal dataset with three modalities 
(a list that consists of 3 numeric matrices). There are six groups of samples within 
the dataset, with ground truth labels given by `example_MultiModalData_labels`. 
Artificial missing values are introduced, leading to an overall missing rate of 20\%.

```{r}
lbls <- example_MultiModalData_labels

str(example_MultiModalData_list)
```

## Model Fit

Then, we fit GPCCA model to the example multi-modal dataset. We specify the size of 
target dimension to be `d = 5`, i.e. the number of latent factors in low-dimensional 
subspace. The default value of ridge regularization parameter `lambda = 0.5` is used. 
Here, we set `verbose = 0` to suppress the printing in every iteration.

```{r}
# GPCCA.fit <- readRDS("D:/Study in UCR/GSR/VK Project 2/R/models/sim/Case A/0.5/MP20/3-4.rds")[[1]]
GPCCA.fit <- GPCCAmodel(X.list = example_MultiModalData_list, d = 5, lambda = 0.5, verbose = 0)
```

The backend of function `GPCCAmodel` is an EM algorithm. Once the model fitting is completed, 
we extract the fitted latent factors, i.e. the joint low-dimensional embeddings of the 
original multi-modal data.

```{r}
LFs <- t(GPCCA.fit$Z)
```

## Downstream Analysis

Next, we attach several more packages for our downstream clustering analysis and visualizations.

```{r, message = FALSE}
library(Seurat)
library(pdfCluster)
library(umap)
library(ggplot2)
```

To obtain the clustering result, we perform Louvain clustering on the joint low-dimensional embeddings learnt by GPCCA. 
The labels of identified clusters are stored in a factor object `grps`.

```{r}
colnames(LFs) <- paste0("LF", 1:ncol(LFs))
rownames(LFs) <- paste0("sample", 1:nrow(LFs))

lc <- FindNeighbors(object = LFs, compute.SNN = TRUE, verbose = FALSE)
lc <- FindClusters(object = lc$snn, verbose = FALSE)

grps <- lc$res.0.8
```

To measure the agreement between the inferred cluster assignment and the true group labels. 
We compute the Adjusted Rand Index (ARI) and print the confusion matrix.

```{r}
adj.rand.index(lbls, grps)

table(lbls, grps)
```

## Visualization

To visualize the fitted latent factors obtained by GPCCA, for example, 
we create a scatterplot of the first two latent factors.

```{r}
par(mfrow = c(1, 2))
plot(x = LFs[, 1], y = LFs[, 2], col = grps,
     type = "p", pch = 16, cex = 0.8, asp = 1,
     xlab = "Latent Factor 1", ylab = "Latent Factor 2",
     main = "Colored by \n identified clusters")
plot(x = LFs[, 1], y = LFs[, 2], col = factor(lbls),
     type = "p", pch = 16, cex = 0.8, asp = 1,
     xlab = "Latent Factor 1", ylab = "Latent Factor 2",
     main = "Colored by \n ground truth labels")
```

Furthermore, to visualize the potential of all joint low-dimensional embeddings combined, 
we create a UMAP based on the reconstructed data representations learnt by GPCCA.

```{r, message = FALSE}
umap_out <- umap(d = LFs)

plotDAT <- data.frame(UMAP1 = umap_out$layout[, 1], UMAP2 = umap_out$layout[, 2],
                      Cluster = as.factor(paste0("cluster ", as.numeric(grps))),
                      Group = as.factor(lbls))

ggplot(plotDAT, aes(x = UMAP1, y = UMAP2, color = Cluster)) + geom_point() + theme_bw() + 
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) + 
  scale_color_brewer(palette = "Set1") + theme(aspect.ratio = 1) + 
  ggtitle("Colored by identified clusters") + 
  theme(plot.title = element_text(hjust = 0.5, size = 16),
        axis.title = element_text(size = 12),
        legend.title = element_text(size = 12)) + 
  guides(color = guide_legend(override.aes = list(size = 5)))

ggplot(plotDAT, aes(x = UMAP1, y = UMAP2, color = Group)) + geom_point() + theme_bw() + 
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) + 
  scale_color_brewer(palette = "Set1") + theme(aspect.ratio = 1) + 
  ggtitle("Colored by ground truth labels") + 
  theme(plot.title = element_text(hjust = 0.5, size = 16),
        axis.title = element_text(size = 12),
        legend.title = element_text(size = 12)) + 
  guides(color = guide_legend(override.aes = list(size = 5)))
```


